---
title: "Sentiment Analysis"
author: "Carolyn Kolaczyk"
date: "2023-06-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
pacman::p_load(dplyr,
               quanteda,
               ggplot2, 
               stringi,
               patchwork) 
```

## State-Level Gay Marriage Legalization

We start by focusing on states that legalized same-sex marriage prior to the Obergefell v. Hodges decision in 2015, which legalized same-sex marriage at the federal level in the United States. 

```{r}
# create function to use with lapply
get_lgbt_data <- function(speech){
  speech = tolower(speech)
  search = "lgbt|trans$|gay|transgender|bisexual|lesbian|queer|asexual|intersex|nonbinary"
  # split into sentences
  sentences<- unlist(stri_split(str = speech, regex = "(?<=[?!.])\\s"))
  # search for lgbt query
  query <- stri_detect_regex(str = sentences, pattern = search) 
  # get lgbt sentences only, combine into 1 document each
  lgbt <- paste(sentences[query], collapse=" ")
  return(lgbt)
}

# read in gay marriage legalization dates
df_marriage <- read.csv('./data/marriage_law_dates.csv')
# convert date from string to date
df_marriage$Date <- as.Date(df_marriage$Date,format='%Y-%m-%d')

# Congressional Record session ids
ids = c('097', '098', '099', '100', '101', '102', '103', '104', '105', 
       '106', '107', '108', '109', '110', '111', '112', '113', '114')

# create new function for splitting data
get_split_data<- function(state_id){
  df_full_made = FALSE # boolean to know when to initialize df_full
  legislation_date = df_marriage[df_marriage$State == state_id,]$Date
  for (session_id in ids){
    #print(paste('Processing session', session_id))
    # read in session data
    file_name = paste('data/processed_data_', session_id,'.csv', sep='')
    data = read.csv(file_name)
    data <- data[,-1]
    # convert date to date type
    data$date <- as.Date(data$date,format='%Y-%m-%d')
    # filter by this state
    state_data <- data[data$state == state_id,]
    # filter so party is only D or R
    state_data <- state_data[state_data['party'] == 'R' | state_data['party'] == 'D',]
    # get only speech, state, party, date
    state_data <- state_data[,c('party', 'state', 'speech', 'date')]
    
    # group by date and party, paste speeches together
    state_data <- state_data %>% 
      group_by(date, party, state) %>% 
      summarise(text = paste(speech, collapse = " "))

    # apply lgbt function to columns
    state_data$text <- unlist(lapply(state_data$text, get_lgbt_data))
    # drop blank columns
    state_data <- state_data[state_data$text != "",]
    # add column with before_after flag
    state_data$before_after <- ifelse(state_data$date <legislation_date, 'before', 'after')
    
    # append dataframes as necessary
    if(dim(state_data)[1] == 0){
      next
    }
    
    if (df_full_made == FALSE){
      df_full = state_data
      df_full_made = TRUE
    }
    else{
      df_full = rbind(df_full, state_data)
    }
  }
  return(df_full)
} 

```

Now we run the above function on all states who legalized gay marriage before the supreme court decision
```{r}
states = df_marriage$State

# iterate over all states
for (state_id in states){
  print(paste('State:', state_id))
  df <- get_split_data(state_id)
  if (state_id == 'CT'){
    df_all_states = df
  }
  else{
    df_all_states = rbind(df_all_states, df)
  }
}

# save to csv
file_name = 'data/state_legalized_marriage_equality_lgbt_data.csv'
write.csv(df_all_states, file_name)
```

### Sentiment Analysis
```{r}
# read in data
file_name = 'data/state_legalized_marriage_equality_lgbt_data.csv'
lgbt_data <- read.csv(file_name)
lgbt_data <- lgbt_data[,-1]
dim(lgbt_data)
head(lgbt_data)
```

We use a dictionary approach for determining sentiment. 

```{r}
# remove apostrophes
lgbt_data$text <- gsub(pattern = "'", "", lgbt_data$text) 
# create corpus
corp <- corpus(lgbt_data, text_field = 'text')

# create dfm, removing punctuation and setting to lower
dfm <- tokens(corp, remove_punct = TRUE) %>% 
  dfm(tolower=TRUE)

# read in positive and negative words (Hu & Liu (2004))
positive_words <- read.csv('data/positive-words.txt', sep='\n', header=FALSE)
negative_words <- read.csv('data/negative-words.txt', sep='\n', header=FALSE)
names(negative_words) <- "word"
names(positive_words) <- "word"
# create lists 
negative_words <- negative_words$word
positive_words <- positive_words$word

# generate sentiment score
# get dfm of only positive words
dfm_positive_words <- dfm %>% dfm_select(positive_words)
# get dfm of only negative words
dfm_negative_words <- dfm %>% dfm_select(negative_words)

# get counts of words (tokens) per document
count_neg_words <- ntoken(dfm_negative_words)
count_pos_words <- ntoken(dfm_positive_words)

# get difference in these counts (# pos - # neg)
sentiment_score <- count_pos_words - count_neg_words

# add this into docvars for dfm_reviews
docvars(dfm, "sentiment_score" ) <- sentiment_score

# create histogram to visualize distribution of sentiment score over all docs
p <- ggplot() + aes(sentiment_score) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Distribution of Sentiment Score") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p

```


Now we compare the distribution of sentiment score across parties, and before and after the legalization of same-sex marriage at the state level. 

```{r}
# D vs R overall
D_scores <- dfm_subset(dfm, party == 'D')$sentiment_score
R_scores <- dfm_subset(dfm, party == 'R')$sentiment_score

p_D <- ggplot() + aes(D_scores) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Distribution of Democrat Sentiment Score") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_R <- ggplot() + aes(R_scores) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Distribution of Republican Sentiment Score") +
  theme(plot.title = element_text(size=15, hjust=0.5))

p_D / p_R 

# before vs after overall
before_scores <- dfm_subset(dfm, before_after == 'before')$sentiment_score
after_scores <- dfm_subset(dfm, before_after == 'after')$sentiment_score
# plot
p_before <- ggplot() + aes(before_scores) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Distribution of Sentiment Score Pre-Legalization") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_after <- ggplot() + aes(after_scores) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Distribution of Overall Sentiment Score Post-Legalization") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_before / p_after


# D vs R vs before vs after (4 plots)
D_before <- dfm_subset(dfm, party == 'D' & before_after=='before')$sentiment_score
D_after <- dfm_subset(dfm, party == 'D' & before_after=='after')$sentiment_score
R_before <- dfm_subset(dfm, party == 'R' & before_after=='before')$sentiment_score
R_after <- dfm_subset(dfm, party == 'R' & before_after=='after')$sentiment_score
# plot
p_D_before <- ggplot() + aes(D_before) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Democrats Pre-Legalization") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_D_after <- ggplot() + aes(D_after) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Democrats Post-Legalization") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_R_before <- ggplot() + aes(R_before) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Republicans Pre-Legalization") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_R_after <- ggplot() + aes(R_after) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Republicans Post-Legalization") +
  theme(plot.title = element_text(size=15, hjust=0.5))

(p_D_before +p_R_before) / (p_D_after+p_R_after)

```

##### Democrats vs Republicans:
We can see that both distributions seem to be centered around zero, but the Democrats distribution seems to be ever so slightly left skewed.

##### Before vs After Legalization:
Both seem to be centered around zero

##### By Party and Date:
We can see that there are much less data points for after legalization. This makes it difficult to do comparison, especially for Republicans. This can also be informative, as it tells us Republicans are discussing lgbtq subjects much less than Democrats. 


Now we compare average sentiment scores.

```{r}
# average sentiment score for each party
print(paste('Avg Sentiment Score for Democrats', mean(D_scores)))
print(paste('Avg Sentiment Score for Republicans', mean(R_scores)))
```
Both scores are negative but the scores for Democrats is more negative.

```{r}
# By party and date
print(paste('Avg Sentiment Score for Democrats Before', mean(D_before)))
print(paste('Avg Sentiment Score for Democrats After', mean(D_after)))
print(paste('Avg Sentiment Score for Republicans Before', mean(R_before)))
print(paste('Avg Sentiment Score for Republicans After', mean(R_after)))
```

We can see that in both cases, the average sentiment scores increases after gay marriage is legalized. It increases by 0.1675 for the Democrats, and 0.3746 for the Republicans. It seems that Republicans sentiment about lgbtq issues increases more after legalization than Democrats. This could make sense if we think about how many/most Democrats already support gay marriage, so it makes sense that their stance wouldn't be as affected by legalization. 

### Extreme Examples
We now look at documents with the highest and lowest sentiment scores, to evaluate the accuracy of our metric.
```{r}
# add sentiment score to corp
docvars(corp) <- docvars(dfm)
# get sentiment scores
scores <- docvars(corp, 'sentiment_score')
scores <- sort(scores)
# get top 5 and bottom 5 values
top_5 <- head(scores)
bottom_5 <- tail(scores)
top_5
bottom_5
```

```{r}
# get most negative docs
print(as.character(corpus_subset(corp, sentiment_score==-51)))
print(as.character(corpus_subset(corp, sentiment_score==-19)))
print(as.character(corpus_subset(corp, sentiment_score==16)))
print(as.character(corpus_subset(corp, sentiment_score==13)))
```
text594: Discussing the murder of private barry winchell, following anti-gay harassment. Criticizing fort campbell and general clark them for not responding better to the harassment before it escalated, and not condemning anti-gay harassment in statements following the murder. 

text718: Talking about how there are no protections for gay students in school against harassment. Talk about student who was killed by a classmate who suspected he was gay. Criticizing schools and teachers and lack of laws in plae protecting gay students. 

text 1712: Speaks in support of allowing gay and lesbian service men/woman to serve in the army, quotes/refers to current service members who also support it

text1383: During pride month, describing how Washington recently became one of 12 states where gay marriage is now legal, and how proud they are of the progress and of the community.


```{r}
# get info about these texts
print(docvars(corpus_subset(corp, sentiment_score==-51)))
print(docvars(corpus_subset(corp, sentiment_score==-19)))
print(docvars(corpus_subset(corp, sentiment_score==16)))
print(docvars(corpus_subset(corp, sentiment_score==13)))
```

Look at the legalization date for these states to see how close these documents are to legalization. 
```{r}
df_marriage[df_marriage$State == 'MA',]$Date
df_marriage[df_marriage$State == 'MN',]$Date
df_marriage[df_marriage$State == 'CA',]$Date
df_marriage[df_marriage$State == 'WA',]$Date
```

None are super close, except for the most negative example, which comes from MA, and is about 6 months before.


## Federal-Level Gay Marriage Legalization

We now look at the states where same-sex marriage was not legal until the federal decision, for comparison. 
```{r}
# function that gets lgbt data but doesn't add before_after
get_data_other_states<- function(state_id){
  df_full_made = FALSE # boolean to know when to initialize df_full
  for (session_id in ids){
    # read in session data
    file_name = paste('data/processed_data_', session_id,'.csv', sep='')
    data = read.csv(file_name)
    # drop doubled index column
    data <- data[,-1]
    # convert data to data type
    data$date <- as.Date(data$date,format='%Y-%m-%d')
    # filter by this state
    state_data <- data[data$state == state_id,]
    # filter so party is only D or R
    state_data <- state_data[state_data['party'] == 'R' | state_data['party'] == 'D',]
    # get only speech, state, party, date
    state_data <- state_data[,c('party', 'state', 'speech', 'date')]

    # group by date and party, paste speeches together
    state_data <- state_data %>% 
      group_by(date, party, state) %>% 
      summarise(text = paste(speech, collapse = " "))
    # apply lgbt function to columns!
    state_data$text <- unlist(lapply(state_data$text, get_lgbt_data))
    # drop blank columns
    state_data <- state_data[state_data$text != "",]
    
    if(dim(state_data)[1] == 0){
      next
    }
    
    if (df_full_made == FALSE){
      df_full = state_data
      df_full_made = TRUE
    }
    else{
      df_full = rbind(df_full, state_data)
    }
  }
  return(df_full)
} 
```

We now apply this function to all states that did not legalize same-sex marriage prior to the Supreme Court decision.

```{r, results='hide'}
states = c('AL', 'AK', 'AZ', 'AR', 'CO',	'FL', 'GA', 'ID', 'IN', 'KS', 'KY', 'LA', 
'MI', 'MS', 'MO', 'MT', 'NE', 'NC', 'ND', 'OH', 'OK', 'OR', 'PA', 'SC', 'SD', 
'TN', 'TX', 'UT', 'WV', 'WI', 'WY')

ids = c('097', '098', '099', '100', '101', '102', '103', '104', '105', 
       '106', '107', '108', '109', '110', '111', '112', '113', '114')

# iterate over all states
df_initialized = FALSE
for (state_id in states){
  print(paste('State:', state_id))
  df <- get_data_other_states(state_id)
  
  # don't save if df is empty
  if(dim(df)[1] == 0){
      print(paste('Nothing for', state_id))
      next
    }
    
  if (df_initialized == FALSE){
    df_all_states = df
    df_initialized  = TRUE
  }
  else{
    df_all_states = rbind(df_all_states, df)
  }
}

# save to csv
file_name = 'data/state_not_legalized_lgbt_data.csv'
write.csv(df_all_states, file_name)
```

### Sentiment Analysis 
```{r}
# read in data
file_name = 'data/state_not_legalized_lgbt_data.csv'
not_legalized_data <- read.csv(file_name)
not_legalized_data <- not_legalized_data[,-1]
dim(not_legalized_data)
head(not_legalized_data)
```

```{r}
# remove apostrophes
not_legalized_data$text <- gsub(pattern = "'", "", not_legalized_data $text) 
# create corpus
corp_2 <- corpus(not_legalized_data, text_field = 'text')

# create dfm, removing punctuation and setting to lower
dfm_2 <- tokens(corp_2, remove_punct = TRUE) %>% 
  dfm(tolower=TRUE)

# read in positive and negative words (Hu & Liu (2004))
positive_words <- read.csv('data/positive-words.txt', sep='\n', header=FALSE)
negative_words <- read.csv('data/negative-words.txt', sep='\n', header=FALSE)
names(negative_words) <- "word"
names(positive_words) <- "word"
# create lists 
negative_words <- negative_words$word
positive_words <- positive_words$word

# generate sentiment score
# get dfm of only positive words
dfm_positive_words <- dfm_2 %>% dfm_select(positive_words)
# get dfm of only negative words
dfm_negative_words <- dfm_2 %>% dfm_select(negative_words)

# get counts of words (tokens) per document
count_neg_words <- ntoken(dfm_negative_words)
count_pos_words <- ntoken(dfm_positive_words)

# get difference in these counts (# pos - # neg)
sentiment_score <- count_pos_words - count_neg_words

# add this into docvars for dfm_reviews
docvars(dfm_2, "sentiment_score" ) <- sentiment_score

# create histogram to visualize distibution of sentiment score over all docs
p <- ggplot() + aes(sentiment_score) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Distribution of Sentiment Score - Other States") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p
```

Like before, the distribution seems to be centered at zero. 

```{r}
# D vs R overall
D_scores <- dfm_subset(dfm_2, party == 'D')$sentiment_score
R_scores <- dfm_subset(dfm_2, party == 'R')$sentiment_score

p_D <- ggplot() + aes(D_scores) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Democrat Sentiment Score - Other States") +
  theme(plot.title = element_text(size=15, hjust=0.5))
p_R <- ggplot() + aes(R_scores) + 
  geom_histogram(binwidth=1, colour="black", fill="white") +
  ggtitle("Republican Sentiment Score - Other States") +
  theme(plot.title = element_text(size=15, hjust=0.5))

p_D / p_R 
```

```{r}
# average sentiment score for each party
print(paste('Avg Sentiment Score for Democrats - other states', mean(D_scores)))
print(paste('Avg Sentiment Score for Republicans - other states', mean(R_scores)))
```
Interestingly, for states that did not legalize gay marriage on their own, we have a very slightly positive average sentiment score for Democrats, and a negative average sentiment score for Republicans.

### Comparison
We now compare the sentiment scores for both groups, states that legalized same-sex marriage prior to the Supreme Court decision, and states that did not.
```{r}
# Prepare for scatterplot
Dem_not_legalize <- docvars(dfm_subset(dfm_2, party == 'D'))[,c('date', 'sentiment_score')]
Rep_not_legalize <- docvars(dfm_subset(dfm_2, party == 'R'))[,c('date', 'sentiment_score')]

# average over date
# nl stands for not legalized
Avg_score_dem_nl <- Dem_not_legalize %>% group_by(date) %>% 
      summarise(avg_score = mean(sentiment_score))
Avg_score_rep_nl <- Rep_not_legalize %>% group_by(date) %>% 
      summarise(avg_score = mean(sentiment_score))
# add columns for party, to use in plotting
Avg_score_dem_nl$party = 'D'
Avg_score_rep_nl$party = 'R'
# combine into one dataframe
avg_score_nl = rbind(Avg_score_dem_nl, Avg_score_rep_nl)
avg_score_nl$date <- as.Date(avg_score_nl$date)

# scatter plot
ggplot(avg_score_nl, aes(x=date, y=avg_score, color=party)) + 
  geom_point() +
  scale_color_manual(values=c('blue','red')) + 
  labs(title="Avg Sentiment Score by Date", 
       subtitle="Non Self-Legalizing States") +
  scale_x_date(breaks = scales::breaks_pretty(10)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),    # Center title position and size
    plot.subtitle = element_text(hjust = 0.5),            # Center subtitle
  )

```


```{r}
# make plot
Dem_l <- docvars(dfm_subset(dfm, party == 'D'))[,c('date', 'sentiment_score')]
Rep_l<- docvars(dfm_subset(dfm, party == 'R'))[,c('date', 'sentiment_score')]

# average over date
# nl stands for not legalized
Avg_score_dem_l <- Dem_l %>% group_by(date) %>% 
      summarise(avg_score = mean(sentiment_score))
Avg_score_rep_l <- Rep_l %>% group_by(date) %>% 
      summarise(avg_score = mean(sentiment_score))
# add columns for party, to use in plotting
Avg_score_dem_l$party = 'D'
Avg_score_rep_l$party = 'R'
# combine into one dataframe
avg_score_l = rbind(Avg_score_dem_l, Avg_score_rep_l)
avg_score_l$date <- as.Date(avg_score_l$date)

# scatter plot
ggplot(avg_score_l, aes(x=date, y=avg_score, color=party)) + 
  geom_point() +
  scale_color_manual(values=c('blue','red')) + 
  labs(title="Avg Sentiment Score by Date", 
       subtitle="States that Legalized Gay Marriage") +
  scale_x_date(breaks = scales::breaks_pretty(10)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),    # Center title position and size
    plot.subtitle = element_text(hjust = 0.5),            # Center subtitle
  )
```

```{r}
# Same plots but on top of each other, for comparison purposes
p1 <- ggplot(avg_score_nl, aes(x=date, y=avg_score, color=party)) + 
  geom_point() +
  scale_color_manual(values=c('blue','red')) + 
  labs(title="Avg Sentiment Score by Date", 
       subtitle="Non Self-Legalizing States") +
  scale_x_date(breaks = scales::breaks_pretty(10)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),    # Center title position and size
    plot.subtitle = element_text(hjust = 0.5),            # Center subtitle
  )

p2 <- ggplot(avg_score_l, aes(x=date, y=avg_score, color=party)) + 
  geom_point() +
  scale_color_manual(values=c('blue','red')) + 
  labs(title="Avg Sentiment Score by Date", 
       subtitle="States that Legalized Gay Marriage") +
  scale_x_date(breaks = scales::breaks_pretty(10)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),    # Center title position and size
    plot.subtitle = element_text(hjust = 0.5),            # Center subtitle
  )

p1 / p2
```

We can see that for states that legalized gay marriage before the supreme court decision, Democrats have a higher range of sentiment scores in the years after 2000. Everything else sees to stay about the same, centered randomly around zero. Although we can also see on the second graph there is a period between about 1987 and 1997 with more negative points than usual. This most likely corresponds with the peak of the AIDS epidemic. 

### Extreme Examples from Other States
```{r}
docvars(corp_2) <- docvars(dfm_2)
# get sentiment scores
scores <- docvars(corp_2, 'sentiment_score')
scores <- sort(scores)
# get top 5 and bottom 5 values
top_5 <- head(scores)
bottom_5 <- tail(scores)
top_5
bottom_5
```

```{r}
# print most neg, 2 most pos
print(as.character(corpus_subset(corp_2, sentiment_score==-17)))
print(as.character(corpus_subset(corp_2, sentiment_score==22)))
print(as.character(corpus_subset(corp_2, sentiment_score==20)))
```

text1796 : Document with the most negative text. This is definitely rated correctly. The speeches in this doc are incredibly anti-trans. They are quoting articles from individuals who regretted their decision, and transgender individuals who are not against bathroom bills. They argue that being transgender is a mental illness and should have stayed classified as such, rather than being changed to gender dysphoria.

text1799 : Multiple speeches in support of the lgbt community, of lgbt congress members, in honor of pride month. This makes sense for a very positive rating. 

text1789 : This document is celebrating victories for the lgbt community. It references the upcoming supreme court decision, and hopes they will legalize gay marriage. We even see some of the same points being made as in the previous text. 

It seems that the most positive documents all occur during pride month. And the negative one is some incredibly anti-trans rhetoric. 

```{r}
# get info about these texts
print(docvars(corpus_subset(corp_2, sentiment_score==-17)))
print(docvars(corpus_subset(corp_2, sentiment_score==22)))
print(docvars(corpus_subset(corp_2, sentiment_score==20)))
```
It is interesting that these are all from Texas. This could suggest that there is a lot of debate regarding the LGBTQ community in Texas. 

